\textbf{The Illusion Game: A Novel Experimental Paradigm Provides Evidence for a General Factor of Visual Illusion Sensitivity and Personality Correlates}

### 2022-11-01 submission/manuscript.tex
citations style

Visual illusions are fascinating stimuli capturing a key feature of our neurocognitive systems. They eloquently show that our brains did not evolve to be perfect perceptual devices providing veridical accounts of physical reality, but integrate prior knowledge and contextual information - blended together in our subjective conscious experience (Carbon, 2014). Despite the long-standing interest within the fields of visual perception (Day, 1972; Eagleman, 2001; Gomez-Villa et al., 2022), consciousness science (Caporuscio et al., 2022; Lamme, 2020), and psychiatry (Gori et al., 2016; Notredame et al., 2014; Razeghi et al., 2022; Teufel et al., 2015), several important issues remain open.

One area of contention concerns the presence of a common mechanism underlying the effect of different illusions (Cretenoud et al., 2020; Hamburger, 2016). While early research has suggested a common factor of illusion sensitivity indexed by overall vision proficiency (Halpern et al., 1999; Thurstone, 1944), recent empirical studies observed at most weak correlations between inter-individual resistance to distinct illusions (Grzeczkowski et al., 2017, 2018). The existence of dispositional correlates of illusion sensitivity has also been controversial, with evidence suggesting a lower illusion sensitivity in patients with schizophrenia and autism (Gori et al., 2016; Grzeczkowski et al., 2018; Notredame et al., 2014; Park et al., 2022; Razeghi et al., 2022), as well as individuals with stronger aggression and narcissism traits (Konrath et al., 2009; Zhang et al., 2017).

Although the nature of the processes underlying illusion perception - whether related to low-level features of the visual processing system (Cretenoud et al., 2019; Gori et al., 2016) or to top-down influences (Caporuscio et al., 2022; Teufel et al., 2018) - remains debated, a growing body of literature proposes to conceptualize illusions under the Bayesian brain hypothesis (Friston, 2010). In this context, illusions are conceptualized as ambiguous percepts (noisy sensory evidence) giving ample weight to prior knowledge to minimize prediction error and provide a coherent perceptual experience. The predictive coding account further provides an explanation regarding the observations from clinical populations. Certain dispositional traits or characteristics (e.g., psychoticism) are seen as driven by alterations in the system's metacognitive components (Adams et al., 2013), resulting in an underweighting of priors during perceptual inferences, and manifesting as a decreased sensitivity to illusions (Koethe et al., 2009).

Despite strong theoretical foundations and hypotheses, the empirical evidence remains scarce, clouded by methodological hurdles. For instance, one key challenge can be found in the difficulty of adapting visual illusions to an experimental setting, which typically requires the controlled modulation of the specific variables of interest. Instead, existing studies typically use only one or a small subset of illusion types, with few contrasting conditions, restricting the findings' generalizability (Bressan \& Kramer, 2021; Cretenoud et al., 2019; Cretenoud et al., 2020). Moreover, conventional paradigms often focus on the participants' subjective experience, by asking them the extent to which they perceive two identical targets as different (Lányi et al., 2022), having them estimate the targets' physical properties (Coren et al., 1976), or having them adjust the targets to perceptually match a reference stimulus (Grzeczkowski et al., 2018; Mylniec \& Bednarek, 2016). This reliance on meta-cognitive judgements about one's subjective experience likely distorts the measurand, limiting the ability to reliably obtain more direct and objective measures of illusion sensitivity (Skottun \& Skoyles, 2014).

To address these issues, we first developed a parametric framework to manipulate visual illusions that we implemented and made accessible in the open-source software \emph{Pyllusion} (Makowski et al., 2021). This software allows us to generate different types of classic visual illusions with a continuous and independent modulation of two parameters: \emph{illusion strength} and \emph{task difficulty} (\textbf{Figure 1}). Indeed, many visual illusions can be seen as being composed of \emph{targets} (e.g., same-length lines), of which perception is biased by the \emph{context} (e.g., in the Müller-Lyer illusion, the same-length line segments appear to have different lengths if they end with inwards vs.~outwards pointing arrows). Past illusion studies traditionally employed paradigms focusing on participants' subjective experience, by asking them the extent to which they perceive two identical targets as different (Lányi et al., 2022), or having them adjust the targets to match a reference stimulus relying only on their perception (Grzeczkowski et al., 2018; Mylniec \& Bednarek, 2016). Alternatively, \emph{Pyllusion} allows the creation of illusions in which the targets are objectively different (e.g., one segment is truly more or less longer than the other), and in which the illusion varies in strength (the biasing angle of the arrows is more or less acute).

@@

The aim of the present preregistered study is three-fold. First, we will test this novel paradigm by investigating if the effect of illusion strength and task difficulty can be manipulated continuously for 10 different classic illusions (Delboeuf, Ebbinghaus, Rod and Frame, Vertical-Horizontal, Zöllner, White, Müller-Lyer, Ponzo, Poggendorff, Contrast). Next, we will investigate the factor structure of illusion-specific performance scores and test the existence of a common latent factor of illusion sensitivity. Finally, we will explore how illusion sensitivity relates to demographic characteristics, contextual variables, and personality traits.

Following open-science standards, all the material (stimuli generation code, experiment code, raw data, analysis script with complementary figures and analyses, preregistration, etc.) is available as \textbf{Supplementary Materials} at \href{https://github.com/RealityBending/IllusionGameValidation}{\textbf{https://github.com/RealityBending/IllusionGameValidation}}.

@@

@@

A pilot study (\emph{n = 46}), of which a full description is available in the Supplementary Materials, was first conducted to determine a sensitive range of stimuli parameters. Then, for each of the 10 illusion types, we generated a total of 134 stimuli. These stimuli resulted from the combination of 15 equally-spaced levels of illusion \emph{strength} (7 negative, i.e., congruent effects; 7 positive, i.e., incongruent effects; and 0) overlapped with 16 non-linearly spaced task \emph{difficulty} levels (i.e., with an exponential, square or cubic spacing depending on the pilot results). For instance, a linear space of {[}0.1, 0.4, 0.7, 1.0{]} can be transformed to an exponential space of {[}0.1, 0.34, 0.64, 1.0{]}, where 0.1 corresponds to the highest difficulty - i.e., the smallest objective difference between targets). For each illusion type, the stimuli were split into two series (56 and 72 stimuli per series) with alternating parameter values to maintain their homogeneity. Additionally, 6 stimuli per illusion type were generated for a practice series using parameters with more extreme variations (i.e., containing very easy trials to help cement the task instructions).

@@

After a brief demographic survey and a practice series of illusions, the first series of 10 illusion blocks was presented in a randomized order, with a further randomization of the stimuli order within each block. Following this first series of blocks, two personality questionnaires were administered, the \emph{IPIP6} (24 items, Sibley et al., 2011) - measuring 6 ``normal'' personality traits (Extraversion, Openness, Conscientiousness, Agreeableness, Neuroticism and Honesty-Humility), and the \emph{PID-5} (25 items, Hopwood et al., 2012) - measuring 5 ``pathological'' personality traits (Disinhibition, Antagonism, Detachment, Negative Affect and Psychoticism). Next, the second series of 10 illusion blocks was presented (with new randomized orders of blocks and trials). In total, each participant underwent 1340 trials of which they had to respond ``as fast as possible without making errors'' (i.e., an explicit double constraint to mitigate the inter-individual variability in the speed-accuracy trade off) by pressing the correct arrow key (left/right, or up/down depending on the illusion type). For instance, in the Müller-Lyer block, participants had to answer which one of the upper or bottom target line was the longest. The task was implemented using \emph{jsPsych} (De Leeuw, 2015), and the set of instructions for each illusion type is available in the experiment code.

@@

Participants were recruited via \emph{Prolific}, a crowd-sourcing platform recognized for providing high quality data (Peer et al., 2022). The only inclusion criterion was a fluent proficiency in English to ensure that the task instructions would be well-understood. Participants were incentivised with a reward of about \textsterling 7.50 for completing the task, which took about 50 minutes to finish. Demographic variables (age, gender, and ethnicity) were self-reported on a voluntary basis.

@@

The first part of the analysis focused on modelling the effect of illusion strength and task difficulty on errors and response time (RT) within each illusion. We started by fitting General Additive Models (GAMs), which can parsimoniously accommodate possible non-linear effects and interactions. Errors were analyzed using Bayesian logistic mixed models, and RTs of correct responses were analyzed using an ex-Gaussian family with the same fixed effects entered for the location \(\mu\) (mean), scale \(\sigma\) (spread) and tail-dominance \(\tau\) of the RT distribution (Balota \& Yap, 2011; Matzke \& Wagenmakers, 2009).

@@

The analysis was carried out using \emph{R 4.2} (R Core Team, 2022), \emph{brms} (Bürkner, 2017), the \emph{tidyverse} (Wickham et al., 2019), and the \emph{easystats} collection of packages (Lüdecke et al., 2021, 2019; Makowski et al., 2020; Makowski, Ben-Shachar, \& Lüdecke, 2019). As the full results are available as supplementary materials, we will focus here on the significant results (based on the Bayes Factor \emph{BF} or the Probability of Direction \emph{pd}, see Makowski, Ben-Shachar, Chen, et al., 2019).

@@

For error rates, most of the models closely matched their GAMs counterpart, with the exception of Delboeuf (for which the GAM suggested a non-monotonic effect of illusion strength with a local minimum at 0) and Zöllner (for which theoretically congruent illusion effects were related to increased error rate). A specific discussion regarding these 2 illusions is available in the Supplementary Materials (Part 1 - Discussion).

@@

Though imperfect, we believe that the random-slope models capture inter-individual differences with more accuracy (and also provide more conservative estimates due to shrinkage) than basic empirical scores, such as the total number of errors, or the average RT. Thus, for each illusion and within each participant, we extracted the effect of illusion strength and its interaction with task difficulty when the illusion effect was incongruent. These twenty participant-level scores were subjected to exploratory factor analysis (EFA). The Method Agreement Procedure (Lüdecke et al., 2020) suggested the presence of 7 latent factors. An oblique (\emph{oblimin} rotation) factor solution explaining 66.69\% of variance suggested separate dimensions for the effect of Zöllner, White, Poggendorff, Contrast, Ebbinghaus, Delboeuf, and a common factor for the parameters related to Müller-Lyer, Vertical-Horizontal, Ponzo and Rod and Frame. We submitted these factors to a second-level analysis and extracted two orthogonal (\emph{varimax} rotation) factors. The first factor was loaded by all the previous dimensions with the exception of Delboeuf, which formed its own separate factor.

@@

The Bayesian t-tests on the effect of sex suggested anecdotal to moderate evidence in favour of the null effect for all scores, with the exception of the sensitivity to the Zöllner illusion, which was higher in males as compared to females (\(\Delta=-0.37\), 95\% CI {[}-0.62, -0.13{]}, \(BF_{10} = 12.74\)). We fitted Bayesian linear models with the education level entered as a monotonic predictor (appropriate for ordinal variables, Bürkner \& Charpentier, 2020), which yielded no significant effects. For age, we fitted two types of models for each score, one general additive models (GAM) and a 2nd order polynomial model. These consistently suggested a significant positive linear relationship between age and Factor \emph{i} (\(pd=100\%\)), as well as the sensitivity to Müller-Lyer (\(pd=100\%\)), Vertical-Horizontal (\(pd=100\%\)), Zöllner (\(pd=100\%\)) and Ebbinghaus (\(pd=99\%\)) illusions (\textbf{Figure 3}).

@@

Regarding ``pathological'' personality traits, the results yielded strong evidence in favor of a negative relationship between illusion scores and multiple traits. \emph{Antagonism} was associated with the sensitivity to Vertical-Horizontal (\(BF_{10} > 100\)), Müller-Lyer (\(BF_{10} = 21.57\)), Ponzo (\(BF_{10} = 17.97\)) illusions, and the Factor \emph{i} (\(BF_{10} = 55.45\)); \emph{Psychoticism} was associated with the sensitivity to Vertical-Horizontal (\(BF_{10} = 66.63\)) and Müller-Lyer (\(BF_{10} = 35.59\)) illusions, and the Factor \emph{i} (\(BF_{10} = 35.02\)); \emph{Disinhibition} was associated with the sensitivity to Vertical-Horizontal (\(BF_{10} = 25.38\)), Zöllner (\(BF_{10} = 7.59\)), Müller-Lyer (\(BF_{10} = 5.89\)) illusions, and the Factor \emph{i} (\(BF_{10} = 31.42\)); and \emph{Negative Affect} was associated with Zöllner (\(BF_{10} = 62.04\)), Vertical-Horizontal (\(BF_{10} = 12.65\)), Müller-Lyer (\(BF_{10} = 3.17\)), and the Factor \emph{i} (\(BF_{10} = 6.39\)). The last remaining trait, \emph{Detachment}, did not share any significant relationship with illusion sensitivity. See Supplementary Materials (Part 2 - Discussion) for a detailed discussion regarding these associations.

@@

This study tested a novel illusion sensitivity task paradigm based on the parametric illusion generation framework (Makowski et al., 2021). Using the carefully generated stimuli in a perceptual decision task, we have shown that a gradual modulation of illusion strength is effectively possible across 10 different types of classic visual illusions. Increasing the illusion strength led to an increase in error likelihood, as well as the average and spread of RTs (but only up to a point, after which participants become faster at responding with the wrong answer). Using mixed models, we were able to statistically quantify the effect of illusions for each illusion and each participant separately. This important methodological step opens the door for new illusions-based paradigms and tasks to study the effect of illusions under different conditions and to measure illusion sensitivity using objective behavioral outcomes - such as accuracy or speed - instead of subjective meta-cognitive reports. This new and complementary approach will hopefully help address some of the longstanding literature gaps, as well as cement illusions as valuable stimuli for the study of cognition.

Our findings suggest that the sensitivity to 10 different types of visual illusions share a common part of variance, supporting the existence of a general factor of illusion sensitivity (Factor \emph{i}). This result comes in a field of mixed findings. In fact, contrary to early studies on visual illusions, more recent research have generally not found any significant evidence for a common stable factor across illusions within individuals (Cretenoud et al., 2019; Cretenoud et al., 2020; Grzeczkowski et al., 2017, 2018; Yang et al., 2012). Instead, past findings suggest illusory effects are highly specific to the perceptual features of the illusions at stake (Cretenoud et al., 2019; Grzeczkowski et al., 2017). It should be noted, however, that most of these studies were low-powered and/or relied on conventional paradigms, such as the adjustment procedure to measure the participants' subjective perception. We believe that our study presents several methodological improvements, including statistical power (high number of trials per participant), homogeneous stimuli (with minimal and highly controlled features) and tasks (decision-making reaction-time task), and a more reliable participant-level score extraction method (based on random-factors models), which in our opinion contributed to the emergence of the common factor.

Finally, we found illusion sensitivity to be positively associated with ``positive'' personality traits, such as agreeableness and honesty-humility, and negatively associated with maladaptive traits such as antagonism, psychoticism, disinhibition, and negative affect. Although the existing evidence investigating links between illusion sensitivity and personality traits is scarce, these results are consistent with past findings relating pathological egocentric beliefs (often associated with psychoticism, Fox, 2006) to reduced context integration, manifesting in a tendency to separate objects from their surroundings when processing visual stimuli (Fox, 2006; Konrath et al., 2009; Ohmann \& Burgmer, 2016). As such, the association between maladaptive traits and lower illusion sensitivity could be linked to a self-centered, decontextualized and disorganized information processing style. Conversely, the relationship between illusion sensitivity and adaptive personality traits is in line with the decreased field dependence (the tendency to rely on external cues in ambiguous contexts) associated with traits negatively correlated with agreeableness and honesty-humility, such as hostility, aggression and narcissism (Konrath et al., 2009; Pessoa et al., 2008; Zhang et al., 2017).

Importantly, these findings highlight the relevance of illusions beyond the field of visual perception, pointing towards an association with high-level domain-general mechanisms. In particular, the evidence in favor of a relationship between maladaptive personality traits and illusion sensitivity is in line with clinical observations, in which a greater resistance to illusions have been reported among patients with schizophrenia (Grzeczkowski et al., 2018; Notredame et al., 2014; Pessoa et al., 2008), especially in association with schizotypal traits such as cognitive disorganization (Cretenoud et al., 2019; Lányi et al., 2022). While the search for the exact mechanism(s) underlying these links is an important goal of future research, our findings unlock the potential of illusion-based tasks as sensitive tools to capture specific inter-individual neuro-cognitive differences.

@@

In conclusion, we strongly invite researchers to explore and re-analyze our dataset with other approaches and methods to push the understanding of visual illusions and illusion sensitivity further. The task, data and analysis script are available in open-access at \href{https://github.com/RealityBending/IllusionGameValidation}{\textbf{https://github.com/RealityBending/IllusionGameValidation}}.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-adams2013computational}{}}%
Adams, R. A., Stephan, K. E., Brown, H. R., Frith, C. D., \& Friston, K. J. (2013). The computational anatomy of psychosis. \emph{Frontiers in Psychiatry}, \emph{4}, 47.

\leavevmode\vadjust pre{\hypertarget{ref-balota2011moving}{}}%
Balota, D. A., \& Yap, M. J. (2011). Moving beyond the mean in studies of mental chronometry: The power of response time distributional analyses. \emph{Current Directions in Psychological Science}, \emph{20}(3), 160--166.

\leavevmode\vadjust pre{\hypertarget{ref-bressan2021most}{}}%
Bressan, P., \& Kramer, P. (2021). Most findings obtained with untimed visual illusions are confounded. \emph{Psychological Science}, \emph{32}(8), 1238--1246.

\leavevmode\vadjust pre{\hypertarget{ref-Burkner2017}{}}%
Bürkner, P.-C. (2017). {brms}: An {R} package for {Bayesian} multilevel models using {Stan}. \emph{Journal of Statistical Software}, \emph{80}(1), 1--28. \url{https://doi.org/10.18637/jss.v080.i01}

\leavevmode\vadjust pre{\hypertarget{ref-burkner2020modelling}{}}%
Bürkner, P.-C., \& Charpentier, E. (2020). Modelling monotonic effects of ordinal predictors in bayesian regression models. \emph{British Journal of Mathematical and Statistical Psychology}, \emph{73}(3), 420--451.

\leavevmode\vadjust pre{\hypertarget{ref-caporuscio2022}{}}%
Caporuscio, C., Fink, S. B., Sterzer, P., \& Martin, J. M. (2022). When seeing is not believing: A mechanistic basis for predictive divergence. \emph{Consciousness and Cognition}, \emph{102}, 103334. \url{https://doi.org/10.1016/j.concog.2022.103334}

>>>>

Visual illusions are fascinating stimuli capturing a key feature of our neurocognitive systems. They eloquently show that our brains did not evolve to be perfect perceptual devices providing veridical accounts of physical reality, but integrate prior knowledge and contextual information - blended together in our subjective conscious experience\textsuperscript{1}. Despite the long-standing interest within the fields of visual perception\textsuperscript{2--4}, consciousness science\textsuperscript{5,6}, and psychiatry\textsuperscript{7--10}, several important issues remain open.

One area of contention concerns the presence of a common mechanism underlying the effect of different illusions\textsuperscript{11,12}. While early research has suggested a common factor of illusion sensitivity indexed by overall vision proficiency\textsuperscript{13,14}, recent empirical studies observed at most weak correlations between inter-individual resistance to distinct illusions\textsuperscript{15,16}. The existence of dispositional correlates of illusion sensitivity has also been controversial, with evidence suggesting a lower illusion sensitivity in patients with schizophrenia and autism\textsuperscript{7--9,16,17}, as well as individuals with stronger aggression and narcissism traits\textsuperscript{18,19}.

Although the nature of the processes underlying illusion perception - whether related to low-level features of the visual processing system\textsuperscript{8,20} or to top-down influences\textsuperscript{5,21} - remains debated, a growing body of literature proposes to conceptualize illusions under the Bayesian brain hypothesis\textsuperscript{22}. In this context, illusions are conceptualized as ambiguous percepts (noisy sensory evidence) giving ample weight to prior knowledge to minimize prediction error and provide a coherent perceptual experience. The predictive coding account further provides an explanation regarding the observations from clinical populations. Certain dispositional traits or characteristics (e.g., psychoticism) are seen as driven by alterations in the system's metacognitive components\textsuperscript{23}, resulting in an underweighting of priors during perceptual inferences, and manifesting as a decreased sensitivity to illusions\textsuperscript{24}.

Despite strong theoretical foundations and hypotheses, the empirical evidence remains scarce, clouded by methodological hurdles. For instance, one key challenge can be found in the difficulty of adapting visual illusions to an experimental setting, which typically requires the controlled modulation of the specific variables of interest. Instead, existing studies typically use only one or a small subset of illusion types, with few contrasting conditions, restricting the findings' generalizability\textsuperscript{12,20,25}. Moreover, conventional paradigms often focus on the participants' subjective experience, by asking them the extent to which they perceive two identical targets as different\textsuperscript{26}, having them estimate the targets' physical properties\textsuperscript{27}, or having them adjust the targets to perceptually match a reference stimulus\textsuperscript{16,28}. This reliance on meta-cognitive judgements about one's subjective experience likely distorts the measurand, limiting the ability to reliably obtain more direct and objective measures of illusion sensitivity\textsuperscript{29}.

To address these issues, we first developed a parametric framework to manipulate visual illusions that we implemented and made accessible in the open-source software \emph{Pyllusion}\textsuperscript{30}. This software allows us to generate different types of classic visual illusions with a continuous and independent modulation of two parameters: \emph{illusion strength} and \emph{task difficulty} (\textbf{Figure 1}). Indeed, many visual illusions can be seen as being composed of \emph{targets} (e.g., same-length lines), of which perception is biased by the \emph{context} (e.g., in the Müller-Lyer illusion, the same-length line segments appear to have different lengths if they end with inwards vs.~outwards pointing arrows). Past illusion studies traditionally employed paradigms focusing on participants' subjective experience, by asking them the extent to which they perceive two identical targets as different\textsuperscript{26}, or having them adjust the targets to match a reference stimulus relying only on their perception\textsuperscript{16,28}. Alternatively, \emph{Pyllusion} allows the creation of illusions in which the targets are objectively different (e.g., one segment is truly more or less longer than the other), and in which the illusion varies in strength (the biasing angle of the arrows is more or less acute).

@@

The aim of the present preregistered (\url{https://osf.io/5d6xp}) study is three-fold. First, we will test this novel paradigm by investigating if the effect of illusion strength and task difficulty can be manipulated continuously for 10 different classic illusions (Delboeuf, Ebbinghaus, Rod and Frame, Vertical-Horizontal, Zöllner, White, Müller-Lyer, Ponzo, Poggendorff, Contrast). Next, we will investigate the factor structure of illusion-specific performance scores and test the existence of a common latent factor of illusion sensitivity. Finally, we will explore how illusion sensitivity relates to demographic characteristics, contextual variables, and personality traits.

@@

\hypertarget{ethics-statement}{%
\subsection{Ethics Statement}\label{ethics-statement}}

This study was approved by the NTU Institutional Review Board (NTU IRB-2022-187) and all procedures performed were in accordance with the ethical standards of the institutional board and with the 1964 Helsinki Declaration. All participants provided their informed consent prior to participation and were incentivized after completing the study.

@@

A pilot study (\emph{n = 46}), of which a full description is available at \href{https://github.com/RealityBending/IllusionGameValidation}{\textbf{https://github.com/RealityBending/IllusionGameValidation}}, was first conducted to determine a sensitive range of stimuli parameters. Then, for each of the 10 illusion types, we generated a total of 134 stimuli. These stimuli resulted from the combination of 15 equally-spaced levels of illusion \emph{strength} (7 negative, i.e., congruent effects; 7 positive, i.e., incongruent effects; and 0) overlapped with 16 non-linearly spaced task \emph{difficulty} levels (i.e., with an exponential, square or cubic spacing depending on the pilot results). For instance, a linear space of {[}0.1, 0.4, 0.7, 1.0{]} can be transformed to an exponential space of {[}0.1, 0.34, 0.64, 1.0{]}, where 0.1 corresponds to the highest difficulty - i.e., the smallest objective difference between targets). For each illusion type, the stimuli were split into two series (56 and 72 stimuli per series) with alternating parameter values to maintain their homogeneity. Additionally, 6 stimuli per illusion type were generated for a practice series using parameters with more extreme variations (i.e., containing very easy trials to help cement the task instructions).

@@

After a brief demographic survey and a practice series of illusions, the first series of 10 illusion blocks was presented in a randomized order, with a further randomization of the stimuli order within each block. Following this first series of blocks, two personality questionnaires were administered, the \emph{IPIP6}\textsuperscript{24 items, 31} - measuring 6 ``normal'' personality traits (Extraversion, Openness, Conscientiousness, Agreeableness, Neuroticism and Honesty-Humility), and the \emph{PID-5}\textsuperscript{25 items, 32} - measuring 5 ``pathological'' personality traits (Disinhibition, Antagonism, Detachment, Negative Affect and Psychoticism). Next, the second series of 10 illusion blocks was presented (with new randomized orders of blocks and trials). In total, each participant underwent 1340 trials of which they had to respond ``as fast as possible without making errors'' (i.e., an explicit double constraint to mitigate the inter-individual variability in the speed-accuracy trade off) by pressing the correct arrow key (left/right, or up/down depending on the illusion type). For instance, in the Müller-Lyer block, participants had to answer which one of the upper or bottom target line was the longest. The task was implemented using \emph{jsPsych}\textsuperscript{33}, and the set of instructions for each illusion type is available in the experiment code.

@@

Participants were recruited via \emph{Prolific}, a crowd-sourcing platform recognized for providing high quality data\textsuperscript{34}. The only inclusion criterion was a fluent proficiency in English to ensure that the task instructions would be well-understood. Participants were incentivised with a reward of about \textsterling 7.50 for completing the task, which took about 50 minutes to finish. Demographic variables (age, gender, and ethnicity) were self-reported on a voluntary basis.

@@

The first part of the analysis focused on modelling the effect of illusion strength and task difficulty on errors and response time (RT) within each illusion. We started by fitting General Additive Models (GAMs), which can parsimoniously accommodate possible non-linear effects and interactions. Errors were analyzed using Bayesian logistic mixed models, and RTs of correct responses were analyzed using an ex-Gaussian family with the same fixed effects entered for the location \(\mu\) (mean), scale \(\sigma\) (spread) and tail-dominance \(\tau\) of the RT distribution\textsuperscript{35,36}.

@@

The analysis was carried out using \emph{R 4.2}\textsuperscript{37}, \emph{brms}\textsuperscript{38}, the \emph{tidyverse}\textsuperscript{39}, and the \emph{easystats} collection of packages\textsuperscript{40--43}. As all the full results have been made available (see \textbf{Data Availability}), we will focus here on the significant results\textsuperscript{based on the Bayes Factor \emph{BF} or the Probability of Direction \emph{pd}, see 44}.

@@

For error rates, most of the models closely matched their GAMs counterpart, with the exception of Delboeuf (for which the GAM suggested a non-monotonic effect of illusion strength with a local minimum at 0) and Zöllner (for which theoretically congruent illusion effects were related to increased error rate). A specific discussion regarding these 2 illusions is available in the study documentation (part 1) at \href{https://github.com/RealityBending/IllusionGameValidation}{\textbf{https://github.com/RealityBending/IllusionGameValidation}}.

@@

Though imperfect, we believe that the random-slope models capture inter-individual differences with more accuracy (and also provide more conservative estimates due to shrinkage) than basic empirical scores, such as the total number of errors, or the average RT. Thus, for each illusion and within each participant, we extracted the effect of illusion strength and its interaction with task difficulty when the illusion effect was incongruent. These twenty participant-level scores were subjected to exploratory factor analysis (EFA). The Method Agreement Procedure\textsuperscript{45} suggested the presence of 7 latent factors. An oblique (\emph{oblimin} rotation) factor solution explaining 66.69\% of variance suggested separate dimensions for the effect of Zöllner, White, Poggendorff, Contrast, Ebbinghaus, Delboeuf, and a common factor for the parameters related to Müller-Lyer, Vertical-Horizontal, Ponzo and Rod and Frame. We submitted these factors to a second-level analysis and extracted two orthogonal (\emph{varimax} rotation) factors. The first factor was loaded by all the previous dimensions with the exception of Delboeuf, which formed its own separate factor.

@@

The Bayesian t-tests on the effect of sex suggested anecdotal to moderate evidence in favour of the null effect for all scores, with the exception of the sensitivity to the Zöllner illusion, which was higher in males as compared to females (\(\Delta=-0.37\), 95\% CI {[}-0.62, -0.13{]}, \(BF_{10} = 12.74\)). We fitted Bayesian linear models with the education level entered as a monotonic predictor\textsuperscript{appropriate for ordinal variables, 46}, which yielded no significant effects. For age, we fitted two types of models for each score, one general additive models (GAM) and a 2nd order polynomial model. These consistently suggested a significant positive linear relationship between age and Factor \emph{i} (\(pd=100\%\)), as well as the sensitivity to Müller-Lyer (\(pd=100\%\)), Vertical-Horizontal (\(pd=100\%\)), Zöllner (\(pd=100\%\)) and Ebbinghaus (\(pd=99\%\)) illusions (\textbf{Figure 3}).

@@

Regarding ``pathological'' personality traits, the results yielded strong evidence in favor of a negative relationship between illusion scores and multiple traits. \emph{Antagonism} was associated with the sensitivity to Vertical-Horizontal (\(BF_{10} > 100\)), Müller-Lyer (\(BF_{10} = 21.57\)), Ponzo (\(BF_{10} = 17.97\)) illusions, and the Factor \emph{i} (\(BF_{10} = 55.45\)); \emph{Psychoticism} was associated with the sensitivity to Vertical-Horizontal (\(BF_{10} = 66.63\)) and Müller-Lyer (\(BF_{10} = 35.59\)) illusions, and the Factor \emph{i} (\(BF_{10} = 35.02\)); \emph{Disinhibition} was associated with the sensitivity to Vertical-Horizontal (\(BF_{10} = 25.38\)), Zöllner (\(BF_{10} = 7.59\)), Müller-Lyer (\(BF_{10} = 5.89\)) illusions, and the Factor \emph{i} (\(BF_{10} = 31.42\)); and \emph{Negative Affect} was associated with Zöllner (\(BF_{10} = 62.04\)), Vertical-Horizontal (\(BF_{10} = 12.65\)), Müller-Lyer (\(BF_{10} = 3.17\)), and the Factor \emph{i} (\(BF_{10} = 6.39\)). The last remaining trait, \emph{Detachment}, did not share any significant relationship with illusion sensitivity.

@@

This study tested a novel illusion sensitivity task paradigm based on the parametric illusion generation framework\textsuperscript{30}. Using the carefully generated stimuli in a perceptual decision task, we have shown that a gradual modulation of illusion strength is effectively possible across 10 different types of classic visual illusions. Increasing the illusion strength led to an increase in error likelihood, as well as the average and spread of RTs (but only up to a point, after which participants become faster at responding with the wrong answer). Using mixed models, we were able to statistically quantify the effect of illusions for each illusion and each participant separately. This important methodological step opens the door for new illusions-based paradigms and tasks to study the effect of illusions under different conditions and to measure illusion sensitivity using objective behavioral outcomes - such as accuracy or speed - instead of subjective meta-cognitive reports. This new and complementary approach will hopefully help address some of the longstanding literature gaps, as well as cement illusions as valuable stimuli for the study of cognition.

Our findings suggest that the sensitivity to 10 different types of visual illusions share a common part of variance, supporting the existence of a general factor of illusion sensitivity (Factor \emph{i}). This result comes in a field of mixed findings. In fact, contrary to early studies on visual illusions, more recent research have generally not found any significant evidence for a common stable factor across illusions within individuals\textsuperscript{12,15,16,20,47}. Instead, past findings suggest illusory effects are highly specific to the perceptual features of the illusions at stake\textsuperscript{15,20}. It should be noted, however, that most of these studies were low-powered and/or relied on conventional paradigms, such as the adjustment procedure to measure the participants' subjective perception. We believe that our study presents several methodological improvements, including statistical power (high number of trials per participant), homogeneous stimuli (with minimal and highly controlled features) and tasks (decision-making reaction-time task), and a more reliable participant-level score extraction method (based on random-factors models), which in our opinion contributed to the emergence of the common factor.

Finally, we found illusion sensitivity to be positively associated with ``positive'' personality traits, such as agreeableness and honesty-humility, and negatively associated with maladaptive traits such as antagonism, psychoticism, disinhibition, and negative affect. Although the existing evidence investigating links between illusion sensitivity and personality traits is scarce, these results are consistent with past findings relating pathological egocentric beliefs\textsuperscript{often associated with psychoticism, 48} to reduced context integration, manifesting in a tendency to separate objects from their surroundings when processing visual stimuli\textsuperscript{19,48,49}. As such, the association between maladaptive traits and lower illusion sensitivity could be linked to a self-centered, decontextualized and disorganized information processing style. Conversely, the relationship between illusion sensitivity and adaptive personality traits is in line with the decreased field dependence (the tendency to rely on external cues in ambiguous contexts) associated with traits negatively correlated with agreeableness and honesty-humility, such as hostility, aggression and narcissism\textsuperscript{18,19,50}.

Importantly, these findings highlight the relevance of illusions beyond the field of visual perception, pointing towards an association with high-level domain-general mechanisms. In particular, the evidence in favor of a relationship between maladaptive personality traits and illusion sensitivity is in line with clinical observations, in which a greater resistance to illusions have been reported among patients with schizophrenia\textsuperscript{7,16,50}, especially in association with schizotypal traits such as cognitive disorganization\textsuperscript{20,26}. While the search for the exact mechanism(s) underlying these links is an important goal of future research, our findings unlock the potential of illusion-based tasks as sensitive tools to capture specific inter-individual neuro-cognitive differences.

@@

\hypertarget{data-availability}{%
\section{Data Availability}\label{data-availability}}

The datasets generated and/or analysed during the current study are available in the GitHub repository \url{https://github.com/RealityBending/IllusionGameValidation}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-carbon2014}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{Carbon, C.-C. \href{https://www.frontiersin.org/articles/10.3389/fnhum.2014.00566}{Understanding human perception by human-made illusions}. \emph{Frontiers in Human Neuroscience} \textbf{8}, (2014).}

\leavevmode\vadjust pre{\hypertarget{ref-day1972}{}}%
\CSLLeftMargin{2. }%
\CSLRightInline{Day, R. H. \href{https://doi.org/10.1126/science.175.4028.1335}{Visual Spatial Illusions: A General Explanation: A wide range of visual illusions, including geometrical distortions, can be explained by a single principle.} \emph{Science} \textbf{175}, 1335--1340 (1972).}

\leavevmode\vadjust pre{\hypertarget{ref-eagleman2001}{}}%
\CSLLeftMargin{3. }%
\CSLRightInline{Eagleman, D. M. \href{https://doi.org/10.1038/35104092}{Visual illusions and neurobiology}. \emph{Nature Reviews Neuroscience} \textbf{2}, 920--926 (2001).}

\leavevmode\vadjust pre{\hypertarget{ref-gomez-villa2022}{}}%
\CSLLeftMargin{4. }%
\CSLRightInline{Gomez-Villa, A., Martín, A., Vazquez-Corral, J., Bertalmío, M. \& Malo, J. \href{https://doi.org/10.1167/jov.22.8.2}{On the synthesis of visual illusions using deep generative models}. \emph{Journal of Vision} \textbf{22}, 2 (2022).}

\leavevmode\vadjust pre{\hypertarget{ref-caporuscio2022}{}}%
\CSLLeftMargin{5. }%
\CSLRightInline{Caporuscio, C., Fink, S. B., Sterzer, P. \& Martin, J. M. \href{https://doi.org/10.1016/j.concog.2022.103334}{When seeing is not believing: A mechanistic basis for predictive divergence}. \emph{Consciousness and Cognition} \textbf{102}, 103334 (2022).}

\leavevmode\vadjust pre{\hypertarget{ref-lamme2020}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{Lamme, V. A. F. \href{https://www.frontiersin.org/articles/10.3389/fpsyg.2020.00083}{Visual functions generating conscious seeing}. \emph{Frontiers in Psychology} \textbf{11}, (2020).}
